Towards the construction of a causal map of the human phenome

### 2017-07-03 docs/manuscript.tex
updated

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Towards the construction of a causal map of the human phenome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{amsmath}

@@

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{26 June 2017}

@@

@@

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant is an
independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses.

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes. Recent developments
in Mendelian randomisation have enabled knowledge of instrumental
variables to be applied using only summary level data (known as
two-sample MR, 2SMR). Here, in order to infer the causal effect of an
exposure on an outcome all that is required is an estimate of the
genetic effects of the instrumenting SNP on the exposure, and the
corresponding estimate of the effect on the outcome. This has two major
advantages. First, GWAS summary data is non-disclosive and often
publicly available. Second, causal inference can be made between
phenotypes even if they have not been measured in the same samples,
limiting the breadth of possible causal estimates only to the
availability of GWAS summary data for the traits in question.

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this problem is not to be understated, and many methods have
been developed that attempt to reliably obtain unbiased causal estimates
under specific models of HP. It is considered best practice to report
estimates from all available methods as sensitivity analyses when
presenting causal estimates, however this strategy is not necessarily
optimal for several reasons. First, if different methods disagree it is
not possible to know which is correct because the appropriate model of
HP is not known. Second, though the IVW approach is most statistically
powerful under no HP, it can have high false negative or low true
positive rates in the presence of HP compared to other methods. Given
that pleiotropy has been hypothesised to be universal, defaulting to the
IVW method in the first instance and using other methods as sensitivity
analyses may not be appropriate. Third, the available methods do not
cover all possible models of HP, and therefore an automated method for
instrument selection may be necessary. Fourth, it could be of interest
to make causal effect estimates for thousands of traits, in which case a
discerning evaluation of each causal effect of interest may not be
possible or convenient.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to
automatically discard genetic variants that are likely to be invalid.
Second, we hypothesised that characteristics of the summary data could
indicate which method would be most reliable, and we introduce new
machine learning approaches that attempt to automate both instrument and
method selection. Using curated GWAS summary data for thousands of
phenotypes, we use these new methods to construct a graph of millions of
causal estimates. We consider this to be only a `first draft' of the
causal map of the human phenome.

@@

The use of summary data in two-sample MR is described in detail
elsewhere. A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained, collecting their effect sizes
and standard errors and effect alleles for the association with the
exposure trait. These can be obtanied manually from complete summary
data or from curated lists of significant GWAS associations. Next, the
effects of those SNPs on the outcome need to be obtained, typically
necessitating complete summary data because it is unlikely that these
SNPs will have reached genome-wide significance (and therefore be
present in curated catalogues). At its most simple implementation, the
regression of the SNP-exposure effect sizes against the SNP-outcome
effect sizes, with greater weight afforded to those SNPs with smaller
SNP-outcome standard errors, provides the estimate of the causal effect
of the exposure on the outcome.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there is sufficient summary data available.
Supplementary table X provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits). Supplementary table X provides a list of the
genetic instruments used for each trait.

@@

\textbf{Mean-based methods:} Here we consider four nested models. The
inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy be directional, where it systematically
occurs in a specific direction. Random effects Egger regression allows
heterogeneity around the directional HP, as long as the HP effects are
not correlated with the SNP-exposure effects (this is the INSIDE
assumption).

The Rucker framework uses estimates of heterogeneity in the IVW and
Egger frameworks to navigate between these nested models. A jackknife
approach (random selection with replacement of instruments) can be used
to obtain a sampling distribution for the model estimate amongst these
four variations. Using 1000 rounds of jackknife estimates, we can obtain
a final estimate using the mean or the median of the distribution. We
only use the jackknife approach for associations where there are 15 or
more variants to avoid combinatorial saturation.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments. This has the advantage that
up to half the instruments can be invalid, and the estimate will remain
unbiased. Developing the approach further to allow stronger instruments
to contribute more towards the estimate can be obtained by obtaining the
median of the weights of each instrument. The penalised weighted median
estimator \ldots{}

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments. This can be
extended in two ways, first by weighting the strength of each cluster by
the inverse of the variance of the SNP-outcome associations in the
cluster; second by weighting the strength of each cluster by the inverse
of the variance of the SNP-exposure associations.

@@

An approach to inferring the causal direction between phenotypes was
developed recently, using the following basic premise. If trait \(A\)
causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Steiger's Z-test of correlated correlations can be used to
formally test the extent to which the two correlations are statistically
different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). In this case the Steiger test applied to each
variant in turn will identify \(g_{A}\) as being unlikely to primarily
associate with \(B\) relative to \(A\). Similarly, for SNPs that
influence confounders of \(A\) and \(B\) or those variants that exhibit
horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant.

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then our
objective is to estimate the variance of the risk liability explained by
the SNP, \(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here,
\(V_e = \pi^2/3\), and \(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the
log odds ratio and \(p\) is the allele frequency of the SNP in the
population. \(p\) can be estimated using the allele frequency of the SNP
in an ascertained sample by deriving the \(2 \times 2\) contingency
table from the odds ratio \(e^\beta\), allele frequency in the
ascertained sample \$p\_\{cc\}, and number of cases \(N_1\) and controls
\(N_0\).

@@

We consider 14 MR methods, for which instruments can be supplied using
two instrument selection strategies, leading to 28 methods in total. In
the context of this analysis, each method is considered to be an
`expert', taking a set of SNP-exposure and SNP-outcome effect sizes and
their standard errors as inputs. Our objective is to select the expert
most likely to be correct for a specific MR analysis.

\paragraph{Mixture of experts}\label{mixture-of-experts}

The mixture of experts (MoE) method is a machine learning approach which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations.

@@

The MoE is trained using datasets generated from simulations. A
\emph{dataset} is the minimum data required to perform 2SMR analysis -
four columns comprising the SNP-exposure and SNP-outcome effects and
standard errors, and rows corresponding to each SNP that is used as an
instrument for the exposure. This \emph{dataset} can be fed into any of
the 28 experts to obtain MR causal effects. The simulations used to
generate these datasets seek to cover a range of pleiotropic scenarios,
including where some proportion of SNPs exhibit directional or balanced
horizontal pleiotropy, or where SNPs influencing confounding variables.

@@

@@

Figure 2a outlines the general strategy behind the MoE implementation.
Briefly, we record 53 metrics about each dataset, hypothesising that we
can use these metrics to predict the performance of each expert, given
some optimisation criteria. For each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics.

Having generated random forest decision trees for each of the 28 methods
using 133,000 of the simulations, we then applied them to the remaining
67,000 datasets to predict which method would have the highest
performance for each of the remaining datasets. Finally we compare the
performance of the method selected by the MoE against all remaining
methods. The default settings for the \texttt{randomForest} package in R
was used to train the models.

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). The random forest for a particular method
\(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
is

\subsubsection{Graph database}\label{graph-database}

@@

For specific hypotheses we strongly recommend that estimates from all
sensitivity analyses are scrutinised and reported.

@@

@@

Number of associations using different approaches Which methods were
preferentially chosen

\subsubsection{A first draft of the causal map of the human
phenome}\label{a-first-draft-of-the-causal-map-of-the-human-phenome}

How many associations reach nominal or FDR corrected significance How
many bi-directional associations How many associations could be declared
for a causal chain of 2

@@

\begin{itemize}
\tightlist
\item
  Limitations of MR that cannot be addressed without biological
  information

  \begin{itemize}
  \tightlist
  \item
    Impossible causal associations
  \item
    Temporal effects
  \item
    Risk liability etc
  \item
    Single variants
  \end{itemize}
\item
  Further developments required to improve causal estimates
\item
  The use of causal chains to improve power
\end{itemize}

>>>>

\documentclass[]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Towards the construction of a causal map of the human phenome},
            pdfauthor={Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden, Tom Gaunt, George Davey Smith},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

@@

\setlength{\droptitle}{-2em}
  \title{Towards the construction of a causal map of the human phenome}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{Gibran Hemani, Jie Zheng, Philip Haycock, Oliver Davies, Jack Bowden,
Tom Gaunt, George Davey Smith}
  \preauthor{\centering\large\emph}
  \postauthor{\par}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{03 July 2017}

\usepackage{amsmath}

@@

\subsection{Abstract}\label{abstract}

A major application for genome-wide association studies (GWAS) has been
the emerging field of causal inference using Mendelian randomisation
(MR), where the causal effect between a pair of traits can be estimated
using only summary level data. MR depends on SNPs exhibiting vertical
pleiotropy, where the SNP influences an outcome phenotype only through
an exposure phenotype. Issues arise when this assumption is violated due
to SNPs exhibiting horizontal pleiotropy, and many methods have been
developed in an attempt to address this. Here we show that the
mechanisms that underlie horizontal pleiotropy are numerous, and that
instrument selection will be increasingly liable to selecting invalid
instruments as GWAS sample sizes continue to grow. We have developed a
mixture of experts machine learning approach (MR-MoE 1.0) that improves
on both power and false discovery rates over all existing methods. Using
the approach, we systematically estimated the causal effects amongst
2407 phenotypes, generating a draft of the causal map of the human
phenome. But we offer caution that its interpretation is far from
straightforward.

@@

Mendelian randomisation (MR) (1,2) exploits genetic pleiotropy to infer
the causal relationships between phenotypes. Suppose that one trait (the
exposure) causally influences another (the outcome). If a SNP influences
the outcome through the exposure then the SNP is exhibiting vertical
pleiotropy. Such a genetic variant is considered to be an instrumental
variable for the exposure, and can be exploited to mimic a randomised
controlled trial, enabling a causal estimate to be made by comparing the
outcome phenotypes between those individuals that have the
exposure-increasing allele against those who do not. Multiple
independent genetic variants for a particular exposure can be used
jointly to improve causal inference, because a) each variant represents
an independent natural experiment, and an overall causal estimate can be
obtained by meta-analysing the single estimates from each instrument;
and b) some violations of the assumptions of MR (most notably that the
SNP influences the outcome through no pathways other than the exposure)
can be evaluated through sensitivity analyses (3--7).

Genome-wide association studies (GWAS) have identified genetic
instrumental variables for thousands of phenotypes (8). Recent
developments in Mendelian randomisation have enabled knowledge of
instrumental variables to be applied using only summary level data
(known as two-sample MR, 2SMR) (9). Here, in order to infer the causal
effect of an exposure on an outcome all that is required is an estimate
of the genetic effects of the instrumenting SNP on the exposure, and the
corresponding estimate of the effect on the outcome. This has two major
advantages. First, GWAS summary data is non-disclosive and often
publicly available. Second, causal inference can be made between
phenotypes even if they have not been measured in the same samples,
limiting the breadth of possible causal estimates only to the
availability of GWAS summary data for the traits in question (10).

Problems with obtaining unbiased causal effects can arise, however, if
the genetic instruments exhibit horizontal pleiotropy (HP), where they
influence the outcome through a pathway other than the exposure. The
extent of this phenomenon is not to be understated, and many methods
have been developed that attempt to reliably obtain unbiased causal
estimates under specific models of HP (3--6). It is considered best
practice to report estimates from all available methods as sensitivity
analyses when presenting causal estimates, however this strategy is not
necessarily optimal for several reasons. First, if different methods
disagree it is not possible to know which is correct because the true
nature of HP exhibited by the instruments is not known. Second, though
the IVW approach is most statistically powerful under no HP, it can have
high false negative or low true positive rates in the presence of HP
compared to other methods. Given that pleiotropy has been hypothesised
to be universal (11,12), defaulting to the IVW method in the first
instance and using other methods as sensitivity analyses may not be
appropriate. Third, the available methods do not cover all possible
models of HP, and therefore an automated method for instrument selection
may be necessary. Fourth, it could be of interest to make causal effect
estimates for thousands of traits, in which case a discerning evaluation
of each causal effect of interest may not be possible or convenient.

In this paper we introduce two innovations towards improving the
reliability of MR estimates. First, we introduce an approach to discard
genetic variants that are likely to be invalid. Second, we hypothesised
that characteristics of the summary data could indicate which method
would be most reliable, and we introduce new machine learning approaches
that attempt to automate both instrument and method selection. Using
curated GWAS summary data for thousands of phenotypes (10), we use these
new methods to construct a graph of millions of causal estimates. We
consider this to be a `first draft' of the causal map of the human
phenome, but raise caution throughout that its interpretation is far
from straightforward

@@

The use of summary data in two-sample MR is described in detail
elsewhere (10). A brief outline of the procedure is as follows. First,
genetic instruments for the exposure trait need to be identified - those
SNPs with \(p < 10^{-8}\) are retained in order to ensure that the first
assumption of Mendelian randomisation (that the instrument associates
with the exposure) is satisfied. We collect their effect sizes, standard
errors and effect alleles for the association with the exposure trait.
These can be obtanied manually from complete summary data or from
curated lists of significant GWAS associations. Next, the effects of
those SNPs on the outcome need to be obtained, typically necessitating
complete summary data because it is unlikely that these SNPs will have
reached genome-wide significance (and therefore be present in curated
catalogues). At its most simple implementation, the regression of the
SNP-exposure effect sizes against the SNP-outcome effect sizes, with
greater weight afforded to those SNPs with smaller SNP-outcome standard
errors, provides the estimate of the causal effect of the exposure on
the outcome. This is known as the inverse-variance weighted (IVW)
method.

Given summary data for a large number of traits, it is straightforward
to exhaustively analyse the causal relationships of every trait against
every other trait for which there is sufficient summary data available.
Supplementary table 1 provides a list of all traits that have available
GWAS summary data, indicating if they have complete summary data (in
which case they can be used as both exposure traits and outcome traits),
or if they only have significant associations (in which they can only be
used as exposure traits).

@@

\textbf{Mean-based methods:} Here we consider four nested models (5).
The inverse variance weighted (IVW) fixed effects meta-analysis approach
assumes that variants exhibit no HP. IVW random effects meta-analysis
relaxes the HP assumption, allowing it to be present but balanced - such
that it only leads to increased heterogeneity around the regression and
not introducing bias. Fixed effects Egger regression (3) relaxes the HP
assumption further by allowing a non-zero intercept which essentially
allows horizontal pleiotropy to be directional, where it systematically
influences the outcome in a specific direction. Random effects Egger
regression allows heterogeneity around the directional HP, as long as
the HP effects are not correlated with the SNP-exposure effects (also
known as the INSIDE assumption).

The Rucker framework (13), adapted to MR (5) uses estimates of
heterogeneity in the IVW and Egger frameworks to navigate between these
nested models. A jackknife approach (random selection with replacement
of instruments) can be used to obtain a sampling distribution for the
model estimate amongst these four variations. Using 1000 rounds of
jackknife estimates, we can obtain a final estimate using the mean or
the median of the distribution. We only use the jackknife approach for
associations where there are 15 or more variants to avoid combinatorial
saturation. The four nested models plus the three Rucker estimates
provide seven mean based estimators.

\textbf{Median-based methods:} An alternative approach is to take the
median effect of all available instruments (4). This has the advantage
that up to half the instruments can be invalid, and the estimate will
remain unbiased. Developing the approach further to allow stronger
instruments to contribute more towards the estimate can be obtained by
obtaining the median of the weights of each instrument. The penalised
weighted median estimator introduces a further weight to the
instruments, penalising any instrument that contributes substantially
towards the heterogeneity statistic. Together, this provides three
median based estimators.

\textbf{Mode-based methods:} Similar to the median, the mode based
estimator clusters the instruments into groups based on similarity of
causal effects, and returns the final causal effect estimate based on
the cluster that has the largest number of instruments (6). This can be
extended in two ways, first by weighting the strength of each cluster by
the inverse of the variance of the SNP-outcome associations in the
cluster; second by weighting the strength of each cluster by the inverse
of the variance of the SNP-exposure associations. When weighting by the
SNP-outcome standard errors we term this a weighted analysis. When
\emph{not} weighting by the exposure standard errors we term this the
`no measuremenent error in the exposure (NOME)' assumption. Together
this provides four mode-based estimators.

@@

An approach to inferring the causal direction between phenotypes was
developed recently (7), using the following basic premise. If trait
\(A\) causes trait \(B\) then

@@

because the \(cor(g_{i}, B)^2 = cor(A, B)^{2} cor(g_{i}, A)^{2}\). This
simple inequality will not hold in some cases, for example
\(\rho_{x, x_o} < \rho_{x,y}\rho_{y,y_o}\) where \(\rho_{x, x_o}\) and
\(\rho_{y, y_o}\) are the precision of the measurements of the \(x\) and
\(y\). Some combinations of confounding effects can also distort the
\(\rho_{g,x}\) and \(\rho_{g,y}\) parameters, as has been discussed in
detail previously. However, we use it here as an inexpensive and
approximate method to identify variants that are likely to be invalid.
Steiger's Z-test of correlated correlations (14) can be used to formally
test the extent to which the two correlations are statistically
different.

Here we adapt this approach to automatically filter SNPs that are liable
to be invalid (Figure 1a). In this case the Steiger test applied to each
variant in turn and we exclude any \(g_{A}\) for which the Steiger
Z-test has \(p > 0.05\), indicating that it is unlikely to primarily
associate with \(B\) relative to \(A\). Similarly, for SNPs that
influence confounders of \(A\) and \(B\) or those variants that exhibit
horizontal pleiotropy, the difference in \(cor(g_{i}, A)^2\) and
\(cor(g_{i}, B)^2\) will be reduced, increasing the likelihood of the
SNP being excluded because the Steiger Z-test is less likely to be
significant.

To estimate \(cor(g, x)^2\), if \(x\) is continuous we obtain the
F-statistic from the reported p-value and sample size and then
\(cor(g, x)^2 = \frac{F}{N - 2 - F}\). If \(x\) is binary then we
estimate the variance of the underlying liability explained by the SNP,
\(cor(g, x)^2 = \frac{V_a}{V_a + V_e}\). Here, \(V_e = \pi^2/3\), and
\(V_a = 2\beta^2p(1-p)\), where \(\beta\) is the log odds ratio and
\(p\) is the allele frequency of the SNP in the population (15). \(p\)
can be estimated using the allele frequency of the SNP in an ascertained
sample by deriving the \(2 \times 2\) contingency table from the odds
ratio \(e^\beta\), allele frequency in the ascertained sample
\$p\_\{cc\}, and number of cases \(N_1\) and controls \(N_0\).

@@

We consider the 14 MR methods described above, for which instruments can
be supplied using two instrument selection strategies, leading to 28
methods in total. In the context of this analysis, each method is
considered to be an `expert', taking a set of SNP-exposure and
SNP-outcome effect sizes and their standard errors as inputs. Our
objective is to select the expert most likely to be correct for a
specific MR analysis.

The mixture of experts method is a machine learning approach (16) which
seeks to divide a parameter space into subdomains, such that a
particular expert is used primarily for problems that reside in a
subdomain most suited to that expert. In this case our objective is to
identify characteristics of the SNP-exposure and SNP-outcome
associations for which one specific MR method is most likely to yield
highest statistical power for non-null associations, and lowest false
discovery rates for null associations. This involves creating a `gating
function', whose purpose is to decide which expert to use for a specific
MR analysis, given the parameter space that is occupied by that dataset.
The metrics are a mixture of regression diagnostics and are described in
Supplementary table 2.

The gating function needs to be trained using data for which the true
causal effect is known, and to this end we generated a large number of
simulated datasets. We trained the gating function using random forest
learning algorithms that seek to identify sectors of the parameter space
that are most likely to return accurate estimates for a particular
expert. Figure 2a illustrates how the gating function is trained using
simulated data. New datasets can then be applied to the trained mixture
of experts (MoE) model (Figure 2b). There are countless ways to
implement this approach, we call this implementation MR-MoE 1.0.

@@

The MoE is trained using datasets generated from simulations (Figure
2a). A \emph{dataset} is the minimum data required to perform 2SMR
analysis - four columns comprising the SNP-exposure and SNP-outcome
effects and standard errors, and rows corresponding to each SNP that is
used as an instrument for the exposure. This \emph{dataset} can be fed
into any of the 28 experts to obtain MR causal effects. The simulations
used to generate these datasets seek to cover a range of pleiotropic
scenarios, including where some proportion of SNPs exhibit directional
or balanced horizontal pleiotropy, or where SNPs influencing confounding
variables.

@@

We aim to maximise statistical power for datasets where
\(\beta_{x,y} \neq 0\) and minimise false discovery rates for datasets
where \(\beta_{x,y} = 0\). To train random forest decision trees to
predict performance for a particular method
\(h(O_{w,d}, \textbf{z}_{d}\) is generated where the training set of
input metrics for dataset \(d\) is \(\textbf{z}_{d}\) and the response
(optimisation function) is

@@

For each training dataset we record a set of 53 metrics
\(\textbf{z}_{d}\) (Supplementary table 2), and an outcome \(O_{w,d}\),
which is a measure of how well that method performed for each particular
simulated dataset. For each of our 28 methods, we need to create a model
that predicts the performance of the method based on metrics generated
from a dataset. To do this, for each method we train random forest
decision trees to predict that method's performance using the dataset's
metrics. The random forest approach is well suited to this problem
because there are likely to be non-continuous combinations of different
metrics that improve on prediction over, for example, a simple linear
model that does not learn about interactions.

A simple hypothetical example - if a dataset has a single outlier but is
otherwise exhibiting no heterogeneity then the following methods could
arguably perform well:

\begin{itemize}
\tightlist
\item
  an IVW fixed effects analysis with the outlier removed, should the
  Steiger test be able to detect the outlier
\item
  a median based approach
\item
  a mode based approach
\end{itemize}

deciding between these methods requires finding, in general, which will
minimise false discovery rates and maximise true positive rates for that
particular scenario.

Having generated random forest decision trees for each of the 28 methods
using 133,000 of the simulations, we then applied them to the remaining
67,000 datasets to predict which method would have the highest
performance for each of the remaining datasets. Finally we compare the
performance of the method selected by the MoE against all remaining
methods. The default settings for the \texttt{randomForest} package in R
(17) was used to train the models. MR-MoE 1.0 is implemented in the
TwoSampleMR R package available at
\href{https://github.com/MRCIEU/TwoSampleMR}{github.com/MRCIEU/TwoSampleMR}.

\subsubsection{Graph database of MR
estimates}\label{graph-database-of-mr-estimates}

@@

For specific hypotheses we strongly recommend that estimates from all
sensitivity analyses are scrutinised and reported. The graph can be
queried directly using the cypher language at
\url{http://shark.epi.bris.ac.uk:7474/browser/} or through a basic web
interface at \url{http://shark.epi.bris.ac.uk:3838}.

@@

As statistical power for GWAS studies improves the likelihood of a
significant association being discovered that doesn't act primarily on
the trait of interest increases (Figure 1a). We evaluated the efficacy
of the Steiger filtering approach for improving instrument selection
using 100,000 simulated datasets comprising both null and non-null
causal models. For each dataset, instruments were selected based on the
tophits strategy and the Steiger filtering strategy, and MR was
performed using 14 different methods based on instruments selected from
each of these strategies.

Figure 1b shows that the tophits strategy led to 49\% of the instruments
being primarily associated with either confounders or the outcome
phenotype, not the exposure phenotype. Applying Steiger filtering
reduced this to 25\%. Consequently, the false discovery rates (FDR) for
12 of the 14 methods reduced substantially when applied using Steiger
filtered instruments (Figure 1c). The true positive rates for the
methods based on Steiger filtering did however reduce slightly for 10 of
the 14 methods.

@@

Following evidence that the Steiger filtering approach can improve on
existing methods, we next hypothesised that a mixture of experts (MoE)
model would be able to use dataset characteristics to predict the most
appropriate of the 28 MR methods and instrument selection strategies
(Figure 2).

The ability to predict the performance for each of these methods is
shown in table 1, with prediction \(R^2\) ranging between 0.04 and 0.24.
The dataset characteristics with the most importance for each of the
predictors differed substantially between each dataset, as well as the
frequency for which each of the methods was selected in the testing
datasets. The FDR of each method when chosen, compared to their averages
across all datasets, reduced; and likewise the true positive rates of
each method when chosen increased compared to their simulation-wide
averages.

We compared the MoE performance in the simulations against each of the
28 methods, testing to see if it outperformed all other single methods.
Figures 3a and 3b show that the MoE approach had the best general
performance, obtaining an AUC of 0.84 in terms of classifying the
simulations as being null or non-null. Notably, the next best methods
were median and mode based estimators using Steiger filtering. Under the
assumption of universal pleiotropy it is clear that all methods suffer
from high false discovery rates.

\subsubsection{Automated MR analysis of 2407
phenotypes}\label{automated-mr-analysis-of-2407-phenotypes}

We applied our analysis using summary data for 2407 phenotypes,
including 149 complex traits and diseases, 575 metabolites (18,19) and
1683 plasma protein levels (20). For the protein levels only the
instruments were available, so they could only be evaluated as exposure
phenotypes. The complex traits and diseases and metabolite levels had
complete summary data available in MR-Base, thus they be evaluated as
both exposures (if they had significant instruments) and outcomes.
Together, we evaluated 715681 relationships. The majority of these
associations could only be evaluated using fewer than 5 SNPs, and so the
Wald ratio or IVW fixed effects methods were used, but for 61029
associations the MR-MoE approach was applied.

There were 5660 associations following Bonferroni correction
(\(p < 7.0\times^{-8}\)). Of these 2918 were obtained from the MR-MoE
analysis, while the remainder were estimated using fewer than 5 SNPs
using the Wald ratio or IVW fixed effects methods.

The frequencies of the methods chosen by the MR-MoE analysis are shown
in Supplementary table 2. Amongst those deemed `significant', the IVW
fixed effects analysis using tophits instruments (the only method
applicable when there is no evidence of horizontal pleiotropy of any
sort) was selected in only 10.4\% of cases.

\subsubsection{Causal associations involving years of
schooling}\label{causal-associations-involving-years-of-schooling}

As an example of the database, we performed a look up of associations
that causally influence years of schooling (21), or that years of
schooling itself causally influences. Using a false-discovery rate of
0.05, 45 traits were returned as having some direction of causality with
years of schooling (Figure 4). Though many of these putative
relationships appear to be plausible, it is clear that there are serious
limitations to using MR as a panacea for causal inference. For example,
bi-directional causal relationships with college completion exist
because of the similarity in trait definitions; and there are several
traits (e.g.~childhood intelligence and birth weight) which appear to be
causally downstream of years of schooling despite this being temporally
impossible. The facility to perform these scans has been automated at
the MR of everything vs everything (MR-EvE) website,
\url{http://shark.epi.bris.ac.uk:3838}, though we urge caution on
interpretation of these findings.

@@

Following a recent flurry in development of MR methodology, we have
devised a machine learning approach, MR-MoE, that seeks to predict the
performance of each MR method, selecting the one which is most likely to
maximise power whilst minimising FDR for a specific dataset. The mixture
of experts approach that we present here is trained using random forest
decision trees applied to extensive and diverse simulations, and we
demonstrate that it makes substantial performance improvements over any
other single method.
